\tit

### 2021-03-01 paper.tex
some translation around All; trim some paragraphs for better page breaks.

\setlength{\abovedisplayskip}{2ex}
\setlength{\belowdisplayskip}{2ex}
\setlength{\abovedisplayshortskip}{0ex}
\setlength{\belowdisplayshortskip}{0ex}

@@

In the definition of {\AB P \AF ☆}, \AB{concat} flattens a list of lists into a single list, defined as a right fold:

\agda{foldr-concat}

@@

In order to convert languages into parsers, it will help to understand how language membership relates to the algebraic structure of lists, specifically the monoid formed by \AIC{[]} and \AF{\_⊙\_}.
Generalizing from languages to functions from lists to \emph{any} type, define
\agda{ν𝒟}
In terms of languages, {\AF{𝒟} \AB{P} \AB{u}} comprises all strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}, i.e., the \AB{u}-suffixes of strings in \AB{P}.

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
\agda{foldl}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), inviting other representations as we will see in \secref{Automatic Differentiation}.

@@

For effective implementations, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient but more complex version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}

@@

}
Moreover, decidability lifts naturally from types to predicates:
\agda{Decidable}
With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
Since this transformation cannot be fully automated, we will instead look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.

@@

To more clearly see the parallel with automatic differentiation (AD), note that AD implementations sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
This work sharing also applies when differentiating languages.
Without this optimization we have
\agda{𝒟′}
The more efficient, work-sharing version:
\agda{ʻ𝒟}
Their equivalence follows from lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
\agda{𝒟′≡ʻ𝒟}
The trie representation accomplishes this sharing by weaving \AF{ν} and \AF{δ} into a single structure based on common prefixes.

@@

The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
First note that there are two categories of operations.
One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
The first category applies to predicates over all types (not just over lists).
Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} apply to all monoids\footnote{The \emph{definitions} of those operations do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).}, while \AF{`} is specific to lists.

\figrefdef{predicates}{Predicate operations}{
\small
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{1.5ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from {\APT{Set} \AB ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{1ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
\setstretch{2.5}
\agda{domain-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).

>>>>

\setlength{\abovedisplayskip}{1.5ex}
\setlength{\belowdisplayskip}{1.5ex}
\setlength{\abovedisplayshortskip}{0ex}
\setlength{\belowdisplayshortskip}{0ex}

@@

In the definition of {\AB P \AF ☆}, \AF{concat} flattens a list of lists into a single list via a right fold.
The predicate {\AD{All} \AB P} holds of a list \AB{ws} when \AB{P} holds for every element of \AB{ws}:\footnote{The data type constructors \AIC{[]} and \AIC{\_∷\_} are overloaded here, referring to both lists (as indices) and \AD{All}.}\\
\agda{foldr-concat}
\vspace{-1ex}
\agda{All}
These three definitions are provided in the Agda standard library \citep{agda-stdlib}.

@@

In order to convert languages into parsers, it will help to understand how language membership relates to the algebraic structure of lists, specifically the monoid formed by \AIC{[]} and \AF{\_⊙\_}.
Generalizing from languages to functions from lists to \emph{any} type, define
\agda{ν𝒟}
For languages, {\AF{𝒟} \AB{P} \AB{u}} is the set of \AB{u}-suffixes from \AB{P}, i.e., the strings \AB{v} such that {\AB{u} \AF ⊙ \AB{v} \AF ∈ \AB{P}}.

Since {\AB{u} \AF ⊙ \AIC{[]} \AD ≡ \AB{u}}, it follows that\footnote{Agda variable names can contain most characters other than spaces and parentheses, e.g., ``\AB{ν∘𝒟}'' as declared here.}\footnote{For functions \AB{f} and \AB{g}, {\AB{f} \AF{≗} \AB{g}} is extensional equality, i.e., {\AS ∀ \AB x \AS → \AB{f} \AB{x} \AD{≡} \AB{g} \AB{x}}.}
\agda{ν∘𝒟}
Moreover, the function \AF{𝒟} distributes over \AIC{[]} and \AF{\_⊙\_}:\footnote{In algebraic terms, argument-flipped \AF{𝒟} is a monoid homomorphism from lists to endofunctions.}
\agda{𝒟[]⊙}
These facts suggest that \AB{𝒟} can be computed one list element at a time via a more specialized operation, as is indeed the case:\footnote{Repeated \AF{δ} application is expressed here via a standard left fold (with {\AB{X} \AS = \AB{A} \AF ✶ \AS → \AB B} in \AF{𝒟foldl}):
\agda{foldl}
}
\agda{δ}
%\vspace{-4ex}
\agda{𝒟foldl}
Each use of \AF{δ} thus takes us one step closer to reducing general language membership to nullability.
%% In other words,%
%% \agda{ν∘foldlδ}
This simple observation is the semantic heart of the syntactic technique of \emph{derivatives of regular expressions} as used for efficient recognition of regular languages \citep{Brzozowski64}, later extended to parsing general context-free languages \citep{Might2010YaccID}.
Lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} liberate this technique from the assumption that languages are represented \emph{symbolically}, by some form of grammar (e.g., regular or context-free), as we will exploit in \secref{Automatic Differentiation}.

@@

For effective implementations, we must bridge the gap between a type (of membership proofs) and its decidable inhabitation.
Fortunately, there is a convenient and compositional way to do so.
For type \AB{A}, the type {\AF{Dec} \AB{A}} contains proof of \AB{A} or a proof of {\AF ¬ \AB{A}} (defined as usual to mean {\AB{A} \AS → \AD{⊥}}):\footnote{This simple \AF{Dec} definition been superceded by a more efficient version \stdlibCitep{Relation.Nullary}.}
\agda{Dec}
For compositionality, we will use a few operations that lift decidability of types to decidability of constructions on those types, as shown in \figrefdef{compositional-dec}{Compositional decidability}{\agda{compositional-dec}}.\footnote{%
A few explanatory remarks:
\begin{itemize}

@@

}
Moreover, decidability lifts naturally from types to predicates:
\agda{Decidable}
With these definitions, we can formulate the problem of decidable language recognition: given a language \AB{P}, construct a term of type {\AF{Decidable} \AB{P}}.
To accomplish this transformation, we will look for decidable building blocks that mirror the predicate vocabulary defined in \secref{Specifying Languages}.

@@

To more clearly see the parallel with automatic differentiation (AD), note that AD implementations sample a function \emph{and} its derivative together to exploit the fact that these two computations typically share much common work.
This work sharing also applies when differentiating languages.
Without this optimization we have
\agda{𝒟′}
The potential work-sharing version:
\agda{ʻ𝒟}
Equivalence follows from lemmas \AF{ν∘𝒟} and \AF{𝒟foldl} of \secref{Decomposing Languages}:
\agda{𝒟′≡ʻ𝒟}
The trie representation exploits this sharing potential by weaving \AF{ν} and \AF{δ} into a single structure based on common prefixes.

@@

The language-building vocabulary defined in \secref{Specifying Languages} can be simplified and generalized.
First note that there are two categories of operations.
One category transforms the codomain (types): \AF{∅}, \AF 𝒰, \AF{\_∪\_}, and \AF{\_∩\_}.
The other transforms the domain (lists): \AF{𝟏}, \AF{\_⋆\_}, \AF{\_☆}, and \AF{`}.
The first category applies to predicates over all types (not just over lists).
Within the second category, \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} can apply to any monoid, while \AF{`} is specific to lists.

\figrefdef{predicates}{Predicate operations}{
\small
\begin{center}
\mathindent-12ex
\agda{Pred}
\end{center}
\vspace{1.5ex}
\hfill
\mathindent0ex
\begin{minipage}[b]{19em}
\agda{codomain-transformers}
\end{minipage}
\hfill
\begin{minipage}[b]{19em}
\agda{domain-transformers}
\end{minipage}
\hfill\;
} shows two parallel collections of predicate operations, each covariantly transforming the codomain (left) or domain (right).\footnote{The types of operations on the left can be further relaxed from {\APT{Set} \AB ℓ} to any codomain type, while keeping \AB{A} fixed, resulting in the usual covariant applicative functor of functions \emph{from} a fixed type \citep{McBride2008APE}.
Dually, the operations on the right (in their current generality) form a covariant applicative functor of functions \emph{to} types.}
These generalized operations then specialize to language operations as in \figrefdef{lang-via-pred}{Languages via predicate operations}{
\mathindent0ex
\begin{center}
\mathindent-12ex
\agda{Lang}
\end{center}
\vspace{1ex}
\hfill
\begin{minipage}[c]{11em}
\setstretch{1.7}
\agda{codomain-ops}
\end{minipage}
\hfill
\begin{minipage}[c]{15em}
\setstretch{2.5}
\agda{domain-ops}
\end{minipage}
\hfill\;
}, which adds language complement (\AF ∁).\footnote{The list-specific operations \AF{\_⊙\_}, \AIC{[]}, and \AF{concat} used in \figref{Lang-ops} have been generalized to monoid operations {\AF{\_∙\_} and \AF{ε}}.
The \emph{definitions} of \AF{𝟏}, \AF{\_⋆\_}, and \AF{\_☆} do not require the monoid properties, but their properties will (\secref{Predicate Algebra}).
}
